---
title: "Anaemia Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Anaemia Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/shahriar26s/anemia-detection-dataset/data\>*

### Reference:

*\<Shahriar26s. (2025). Anemia Detection Dataset. Kaggle. https://www.kaggle.com/datasets/shahriar26s/anemia-detection-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r Load dataset}
# Load anaemia dataset with appropriate column types
AnaemiaData <- read.csv("anemia_dataset.csv", colClasses = c(
  Red_Pixel = "numeric",
  Green_pixel = "numeric",
  Blue_pixel = "numeric",
  Hb = "numeric",           # Haemoglobin levels
  Anaemic = "factor"        # Yes or No (classification target)
))

# View the structure of the dataset
str(AnaemiaData)

# Quick check on data
summary(AnaemiaData)

#View dataset
View(AnaemiaData)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Frequency of the target variable
table(AnaemiaData$Anaemic)

# Frequency of Hb ranges (e.g. grouping Hb levels)
cut_Hb <- cut(AnaemiaData$Hb, breaks = c(0, 7, 10, 13, 16, Inf), right = FALSE)
table(cut_Hb)

```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean
mean(AnaemiaData$Red_Pixel)
mean(AnaemiaData$Green_pixel)
mean(AnaemiaData$Blue_pixel)
mean(AnaemiaData$Hb)

# Median
median(AnaemiaData$Red_Pixel)
median(AnaemiaData$Green_pixel)
median(AnaemiaData$Blue_pixel)
median(AnaemiaData$Hb)

# Mode function
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

get_mode(AnaemiaData$Anaemic)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Range
range(AnaemiaData$Hb)

# Variance and standard deviation
var(AnaemiaData$Hb)
sd(AnaemiaData$Hb)

# Skewness and kurtosis (requires e1071)
library(e1071)
skewness(AnaemiaData$Hb)
kurtosis(AnaemiaData$Hb)

```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation matrix among pixel values and Hb
cor(AnaemiaData[, c("Red_Pixel", "Green_pixel", "Blue_pixel", "Hb")])

# Chi-square test between Anaemic and binned Hb
Hb_group <- cut(AnaemiaData$Hb, breaks = c(0, 10, 12, Inf), labels = c("Low", "Medium", "High"))
chisq.test(table(Hb_group, AnaemiaData$Anaemic))
```

## ANOVA
```{r ANOVA}
# One-way ANOVA: Does Hb significantly differ by Anaemic status?
anova_result <- aov(Hb ~ Anaemic, data = AnaemiaData)

# Summary of ANOVA test
summary(anova_result)

# Tukey HSD post-hoc test
TukeyHSD(anova_result)
```

## Plots
```{r Plots}
# Histogram for Hb
hist(AnaemiaData$Hb, main = "Distribution of Hb Levels", xlab = "Hb", col = "skyblue", border = "white")

# Bar plot for Anaemic status
barplot(table(AnaemiaData$Anaemic), main = "Anaemic Status Frequency", col = c("red", "green"), ylab = "Count")

# Boxplots for pixel values
boxplot(AnaemiaData$Red_Pixel, main = "Red Pixel Distribution", col = "tomato")
boxplot(AnaemiaData$Green_pixel, main = "Green Pixel Distribution", col = "lightgreen")
boxplot(AnaemiaData$Blue_pixel, main = "Blue Pixel Distribution", col = "lightblue")

# Boxplot of Hb by Anaemic status
boxplot(Hb ~ Anaemic, data = AnaemiaData,
        main = "Hb Levels by Anaemic Status", col = c("red", "green"),
        ylab = "Haemoglobin (Hb)")

# Scatterplot matrix for pixel values and Hb
pairs(AnaemiaData[, c("Red_Pixel", "Green_pixel", "Blue_pixel", "Hb")],
      main = "Scatterplot Matrix", col = AnaemiaData$Anaemic)

# Colored scatterplot: Hb vs Red_Pixel
plot(AnaemiaData$Red_Pixel, AnaemiaData$Hb,
     col = ifelse(AnaemiaData$Anaemic == "Yes", "red", "green"),
     pch = 19, xlab = "Red Pixel", ylab = "Hb", main = "Hb vs Red Pixel")

legend("topright", legend = c("Anaemic", "Non-Anaemic"),
       col = c("red", "green"), pch = 19)

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check if any missing values exist
any(is.na(AnaemiaData))  # Returns TRUE if any NA is present

# Count of missing values per column
colSums(is.na(AnaemiaData))

# Summary to visually inspect data completeness
summary(AnaemiaData)
```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load required library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Create index for 70% training data
train_index <- createDataPartition(AnaemiaData$Anaemic, p = 0.7, list = FALSE)

# Split the data
train_data <- AnaemiaData[train_index, ]
test_data  <- AnaemiaData[-train_index, ]

# Confirm dimensions
cat("Training set rows:", nrow(train_data), "\n")
cat("Testing set rows:", nrow(test_data), "\n")
```

## Bootstrapping
```{r Boot}
# Load required libraries
library(caret)

# Set seed for reproducibility
set.seed(123)

# Define bootstrap control
bootstrap_control <- trainControl(
  method = "boot",
  number = 100,              # 100 bootstrap resamples
  savePredictions = "final",
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

# Make sure the response variable is a factor with valid levels
train_data$Anaemic <- factor(train_data$Anaemic, levels = c("Yes", "No"))

# Train a logistic regression model with bootstrapping
boot_model <- train(
  Anaemic ~ ., data = train_data,
  method = "glm",
  family = "binomial",
  trControl = bootstrap_control,
  metric = "ROC"
)

# Output model summary
print(boot_model)
```

## Cross-validation
```{r Cross-validation}
# Load required library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Define 10-fold cross-validation control
cv_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Ensure the response variable is a factor with correct levels
train_data$Anaemic <- factor(train_data$Anaemic, levels = c("Yes", "No"))

# Train a logistic regression model using cross-validation
cv_model <- train(
  Anaemic ~ ., data = train_data,
  method = "glm",
  family = "binomial",
  trControl = cv_control,
  metric = "ROC"
)

# Print model summary
print(cv_model)

```

## Training Different Models
```{r Different Models}
# Load required packages
library(caret)
library(gbm)
library(randomForest)

# Set seed for reproducibility
set.seed(123)

# Ensure target is a factor
train_data$Anaemic <- factor(train_data$Anaemic, levels = c("Yes", "No"))

# Define common CV control
cv_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Train Logistic Regression
model_glm <- train(
  Anaemic ~ ., data = train_data,
  method = "glm",
  family = "binomial",
  trControl = cv_control,
  metric = "ROC"
)

# Train Random Forest
model_rf <- train(
  Anaemic ~ ., data = train_data,
  method = "rf",
  trControl = cv_control,
  metric = "ROC"
)

# Train Gradient Boosting Machine
model_gbm <- train(
  Anaemic ~ ., data = train_data,
  method = "gbm",
  verbose = FALSE,
  trControl = cv_control,
  metric = "ROC"
)

# Print summaries
print(model_glm)
print(model_rf)
print(model_gbm)
```

## Performance Comparison
```{r Performance Comparison}
# Compare the three models
results <- resamples(list(
  GLM = model_glm,
  RF = model_rf,
  GBM = model_gbm
))

# Summary statistics (ROC, Sensitivity, Specificity)
summary(results)

# Boxplots for comparison
bwplot(results, metric = "ROC")
bwplot(results, metric = "Sens")
bwplot(results, metric = "Spec")

# Dotplot comparison
dotplot(results, metric = "ROC")
```

## Saving Model
```{r Saving Model}
# Load the saved GBM model (when needed)
loaded_gbm_model <- readRDS("./models/saved_gbm_model.rds")

# Example of new data for prediction (adjust this to match your dataset)
new_data <- data.frame(
  Red_Pixel = 45.0,
  Green_pixel = 28.5,
  Blue_pixel = 26.0,
  Hb = 10.5
)

# Use the loaded GBM model to make predictions
predictions_loaded_model <- predict(loaded_gbm_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

